### 2022-11-10 NSF2022/Description/specificaim2.tex
working on SA2

\subsection{Specific Aim 2: Efficient, high-order methods for
large-scale simulations}
\label{sec:specificaim2}
\begin{itemize}
%  \item Make clear that we know how to address NSI and fast summation
  \item Talk about collision as per talk with Bryce
%  \item Wen's paper can be used to give numbers on how large or a
%    problem could hypothetically be considered
%  \item Emphasize that we've learned a lot from just $\mathcal(O)(100)$
%    particles, but now it's time to do more
  \item Confinement and rheology
\end{itemize}

% ----------------------------------------------------------------------
Over the past decade, there has been an explosion of interest in
small-scale processes that utilize capillary forces and van der Waals
interactions to coordinate movement and bind microscopic components in
solvent~\cite{Pandey2011, Zhang2017, Siontorou2017}. Additionally,
hydrodynamic interactions are central to fabricate complex,
three-dimensional microstructures~\cite{Dasgupta2017, Leong2007,
Reynolds2019, Cho2010}. Hydrodynamic effects cannot be ignored since
they, like for the rates of biological functions like pore
dynamics~\cite{RYHAM20112929}, set the rate of dissipation. In Specific
Aim 2, we propose novel and efficient high-order numerical methods to
simulate dynamic assembly of particles under the HAP in a viscous
solvent. As described in~\eqref{eqn:stokes} and illustrated in
Figure~\ref{fig:flow_map}, the solvent phase is modeled by the Stokes
equations for an incompressible, viscous fluid, and these equations are
coupled to the non-linear screened-Laplace equation~\eqref{eqn:phase}
through viscous and hydrophobic stress balance.

%\label{subsec:specific_aim_2}
%\begin{wrapfigure}[14]{r}{0.61\textwidth}
%  \centering{\includegraphics[width=0.60\textwidth]{Figures/domain.pdf}}
%    \vspace{-15pt}
%    \caption{\label{fig:flow_map} \footnotesize A collection of 3D
%    amphiphilic particles suspended in a solvent. The solvent velocity
%    satisfies the Stokes equations, and the hydrophobic forces depend on
%    the solvent activity that satisfies the screened Laplace equation.
%    The geometry is updated by solving the mobility problem.} 
%\end{wrapfigure}
The HAP model requires solving exterior problems for the order
parameter~\eqref{eqn:phase} and the Stokes equations~\eqref{eqn:stokes}.
We will develop methods based on combinations of volume and layer
potentials, and inexact Newton's method. As mentioned in
Section~\ref{sec:bie}, numerical methods to solve these integral
equations (IEs) have several advantages over their PDE-based
counterpart. The PIs have demonstrated that IEs are a powerful tool to
simulate two-dimensional suspensions of amphiphilic
particles~\cite{Fu2018_SIAM, FuQuRyYo22, fu-ryh-qua-you2022}, and
elements of these formulations have been developed in three
dimensions~\cite{ying_2006, manasthesis, rac-gre2016}. This proposal
will extend the two-dimensional methods to more general amphiphilic
particle models and also to three dimensions. In addition, the methods
will allow us to use results we have shown for $\mathcal{O}(100)$
granules, and extend them to suspensions with 100X--1000X more granules.
These methods will be motivated by models and analysis developed in
Specific Aims 1 and 3. \S\ref{subsec:AC} describes how the non-linear
equation for the order parameter~\eqref{eqn:phase} will be iteratively
solved using an IE formulation. Then, \S\ref{subsec:NumericalIssues}
describes how we will combine accurate quadrature and fast summation
methods to accelerate the time to accurate solutions. A natural
extension to create simulations that are realizable by experimentalists
is to include confinement, and this is discussed in
\S\ref{subsec:Confinement}. This extension will also allow us to develop
new rheological properties of the suspensions.  Finally,
\S\ref{subsec:timeStepping} proposes collision-aware adaptive time
stepping methods.

% ----------------------------------------------------------------------
\subsubsection{An inexact Newton solver for the phase transition PDE}
\label{subsec:AC}

\begin{wrapfigure}[28]{l}{1.8in}
  \vspace{-8pt}
  \centering
  \input{figures/doublewell.tikz} 
  \caption{\label{fig:CA} \footnotesize (a) A double well potential. (b)
  The solution of the one-dimensional phase
  transition~\eqref{eqn:phase}. (c) The convergence of exact and inexact
  Newton's methods.}
\end{wrapfigure}
The PIs have developed numerical methods to solve~\eqref{eqn:phase} when
$W(\phi) = \phi^2$---this choice results in the linear screened Laplace
equation. For a general double well potential $W$ with local minima at
$\phi_0$ and $\phi_1$, we propose to interpret the solution as the root
finding problem for
\begin{align}
  \label{eqn:phaseRoot}
  J[\phi] = -\epsilon^2 \Delta \phi + \tfrac{1}{2}W'(\phi).
\end{align}
Equation~\eqref{eqn:phaseRoot} will be solved with Newton's method,
$\phi^{N+1} = \phi^{N} - \alpha (dJ^N)^{-1} J(\phi^N)$, where $dJ^N =
-\epsilon^2 \Delta + \tfrac{1}{2}W''(\phi^N)$ is the Fr\'{e}chet
derivative of $F$, and $\alpha$ is the step size that will be determined
using backtracking. This requires solving the variable coefficient PDE
$-\epsilon^2 \Delta v + \tfrac{1}{2}W''(\phi^{N}) v = J[\phi^N]$, which
is challenging with an integral equation since a fundamental solution is
not readily available. However, by choosing a double well potential with
$W''(\phi_0) = W''(\phi_1) = \beta$, we propose to instead use an
inexact Newton's method where $dJ^{-1}$ is approximated with the
solution of $-\epsilon^2 \Delta v + \tfrac{\beta}{2} v = J[\phi^N]$. We
test the efficacy of the exact and inexact Newton's method applied
to~\eqref{eqn:phaseRoot} in $(0,1) \subset \RR$ by discretizing $dJ$
with a second-order finite difference stencil. The double well potential
is illustrated in Figure~\ref{fig:CA}(a), and both the exact and inexact
Newton's methods converge to the solution in Figure~\ref{fig:CA}(b).
Figure~\ref{fig:CA}(c) shows the norm of the residual at each Newton
iterate when using the exact second derivative (red curve) and the
approximate second derivative (blue curve). While the inexact Newton's
method requires roughly twice as many iterations to reach of a tolerance
of $10^{-4}$, each iteration is significantly cheaper to apply.

Having demonstrated that the inexact Newton method converges in one
dimension, it will be coupled with IE formulations to
solve~\eqref{eqn:phase} in two and three dimensions. The solution of
$-\epsilon^2 \Delta v + \tfrac{\beta}{2} v = J[\phi^N]$ can be written
as $v = v^P + v^H$ where $v^P$ is a volume potential and $v^H$ is a
homogeneous solution of the PDE, which is represented as a layer
potential, and is used to satisfy the boundary condition. The volume
potential is
\begin{align*}
  v^{P}(\xx) = \int_{\Omega(t)} G(\xx,\yy) J[\phi^N](\yy)\, d\yy, 
    \quad \xx \in \Omega(t),
\end{align*}
where $G$ is the fundamental solution of $-\epsilon^2 \Delta +
\tfrac{\beta}{2}$. Traditionally, volume potentials are more challenging
to compute than layer potentials, but new methods have recently been
developed that allows them to be computed accurately and
efficiently~\cite{fry-leh-tor2018, and-zhu-vee2022}. This approach of
combining volume and layer potentials has successfully been used by PI
BQ~\cite{kro-qua2010} and others~\cite{fry-kro-tor2019, kli-ask-kro2020,
che-hua-lei2005}.

% ----------------------------------------------------------------------
\subsubsection{Increased accuracy and reduced CPU time}
\label{subsec:NumericalIssues}
While IE formulations of PDEs have several advantages, several
challenges must be overcome before the code can be used to simulate
systems with large numbers of granules. For example, when granules are
close to one another, which is inevitable in dense suspensions, the
resulting integrals are nearly-singular. This issue is well-known in the
IE community and several algorithms have been developed in the last 30
years. An incomplete list of methods that address near-singular
integration include panel-based methods~\cite{helsing_2008a,
helsing_tutorial_2012}, Quadrature by Expansions
(QBX)~\cite{din-hua-mar-tan2021, klo-bar-gre-one2013, ste-bar2022,
af2018sisc}, and interpolation-based methods~\cite{ying_2006,
qua-bir2014, rah-vee-zor-bir2015}. The proposed research, which is
described next,  will combine several methods including singularity
subtraction, analytic function theory, and spherical basis
representations.

@@

% ----------------------------------------------------------------------
\subsubsection{Incorporating confinement}
\label{subsec:Confinement}
In experiments, it is known that interactions between bounding walls and
granules cannot be ignored~\cite{qua-gan-you2021,
KaouiHartingMisbah2011_PRE, her-sto-gra2005}. Therefore, the necessary
modifications to the code will be performed so that confinement can be
considered. PI BQ has considered suspensions with confinement for
several projects including dynamic suspensions~\cite{qua-bir2014,
bys-sha-qua2020} and porous media~\cite{chi-moo-qua2020,
dea-qua-bir-jua2018, moo-che-chi-qua2022, qua-gan-you2021}.

>>>>

\subsection{Specific Aim 2: Efficient, high-order methods for
large-scale simulations}
\label{sec:specificaim2}
% ----------------------------------------------------------------------
Over the past decade, there has been an explosion of interest in
small-scale processes that utilize capillary forces and van der Waals
interactions to coordinate movement and bind microscopic components in
solvent~\cite{Pandey2011, Zhang2017, Siontorou2017}. Additionally,
hydrodynamic interactions are central to fabricate complex,
three-dimensional microstructures~\cite{Dasgupta2017, Leong2007,
Reynolds2019, Cho2010}. Hydrodynamic effects cannot be ignored since
they, like for the rates of biological functions like pore
dynamics~\cite{RYHAM20112929}, set the rate of dissipation. The success
of Specific Aims 1 and 3 depends on the ability to accurately and
efficiently simulate dense suspensions of granules governed by
equations~\eqref{eq:RBT}--\eqref{eqn:stressbalance}. Because of their
ability to resolve complex unbounded domains and achieve high-order
accuracy, integral equation methods will be a central theme of the
computational tools. As discussed in Section~\ref{sec:bie}, numerical
methods to solve IEs have several advantages over their PDE-based
counterpart. The PIs have demonstrated that IEs are a powerful tool to
simulate two-dimensional suspensions of amphiphilic
particles~\cite{Fu2018_SIAM, FuQuRyYo22, fu-ryh-qua-you2022}, and
elements of these formulations have been developed in three
dimensions~\cite{ying_2006, manasthesis, rac-gre2016}. In our current
implementation, we can simulate $\mathcal{O}(100)$ two-dimensional
granules, and the proposed research will allow for the suspensions with
100X--1000X more two-dimensional granules and $\mathcal{O}(1000)$
three-dimensional granules.

\begin{wrapfigure}[27]{l}{1.8in}
  \vspace{-8pt}
  \centering
  \input{figures/doublewell.tikz} 
  \caption{\label{fig:CA} \footnotesize (a) A double well potential. (b)
  The solution of the one-dimensional phase
  transition~\eqref{eqn:phase}. (c) The convergence of exact and inexact
  Newton's methods.}
\end{wrapfigure}

The order parameter is governed by the PDE~\eqref{eqn:phase} in the
solvent phase $\Omega(t)$. For the single-well $W(\phi) = \phi^2$,
equation~\eqref{eqn:phase} is linear, and the PIs have demonstrated that
IEs are a powerful tool to solve for the order
parameter~\cite{Fu2018_SIAM, FuQuRyYo22, fu-ryh-qua-you2022}. However,
for a general double-well potential $W(\phi)$, the resulting PDE is
nonlinear, motivating the following problem:

% ----------------------------------------------------------------------
\begin{quotation}
  \noindent
  \textbf{Problem 3.} How can IEs be coupled with a Newton solver to
  solve the non-linear PDE~\eqref{eqn:phase}?
\end{quotation}

The PIs have developed numerical methods to solve~\eqref{eqn:phase} when
$W(\phi) = \phi^2$---this choice results in the linear screened Laplace
equation. For a general double well potential $W$ with local minima at
$\phi_0$ and $\phi_1$, we propose to interpret the solution as the root
finding problem for
\begin{align}
  \label{eqn:phaseRoot}
  F[\phi] = -\epsilon^2 \Delta \phi + \tfrac{1}{2}W'(\phi).
\end{align}
Equation~\eqref{eqn:phaseRoot} will be solved with Newton's method,
$\phi^{N+1} = \phi^{N} - \alpha (dF^N)^{-1} F(\phi^N)$, where $dF^N =
-\epsilon^2 \Delta + \tfrac{1}{2}W''(\phi^N)$ is the Fr\'{e}chet
derivative of $F$, and $\alpha$ is the step size that will be determined
using backtracking. This requires solving the variable coefficient PDE
$-\epsilon^2 \Delta v + \tfrac{1}{2}W''(\phi^{N}) v = F[\phi^N]$, which
is challenging with an integral equation since a fundamental solution is
not readily available. However, by choosing a double well potential with
$W''(\phi_0) = W''(\phi_1) = \beta$, we propose to instead use an
inexact Newton's method where $dF^{-1}$ is approximated with the
solution of $-\epsilon^2 \Delta v + \tfrac{\beta}{2} v = F[\phi^N]$. We
test the efficacy of the exact and inexact Newton's method applied
to~\eqref{eqn:phaseRoot} in $(0,1) \subset \RR$ by discretizing $dF$
with a second-order finite difference stencil. The double well potential
is illustrated in Figure~\ref{fig:CA}(a), and both the exact and inexact
Newton's methods converge to the solution in Figure~\ref{fig:CA}(b).
Figure~\ref{fig:CA}(c) shows the norm of the residual at each Newton
iterate when using the exact second derivative (red curve) and the
approximate second derivative (blue curve). While the inexact Newton's
method requires roughly twice as many iterations to reach of a tolerance
of $10^{-4}$, each iteration is significantly cheaper to apply.

Having demonstrated that the inexact Newton method converges in one
dimension, it will be coupled with IE formulations to
solve~\eqref{eqn:phase} in two and three dimensions. The solution of
$-\epsilon^2 \Delta v + \tfrac{\beta}{2} v = F[\phi^N]$ can be written
as $v = v^P + v^H$ where $v^P$ is a volume potential and $v^H$ is a
homogeneous solution of the PDE, which is represented as a layer
potential, and is used to satisfy the boundary condition. The volume
potential is
\begin{align*}
  v^{P}(\xx) = \int_{\Omega(t)} G(\xx,\yy) F[\phi^N](\yy)\, d\yy, 
    \quad \xx \in \Omega(t),
\end{align*}
where $G$ is the fundamental solution of $-\epsilon^2 \Delta +
\tfrac{\beta}{2}$. Traditionally, volume potentials are more challenging
to compute than layer potentials, but new methods have recently been
developed that allows them to be computed accurately and
efficiently~\cite{fry-leh-tor2018, and-zhu-vee2022}. This approach of
combining volume and layer potentials has successfully been used by PI
BQ~\cite{kro-qua2010} and others~\cite{fry-kro-tor2019, kli-ask-kro2020,
che-hua-lei2005}.
% ----------------------------------------------------------------------
\begin{quotation}
  \noindent
  \textbf{Problem 4.} How can novel quadrature, time adaptivity, fast
  summation, and collision-detection be combined to accurately and
  efficiently solve the governing
  equations~\eqref{eq:RBT}--\eqref{eqn:stressbalance} in bounded and
  unbounded domains in two and three dimensions?
\end{quotation}

\begin{wrapfigure}[]{r}{2.1in}
  \centering
  \input{figures/near_sing.tikz}
  \caption{\label{fig:interpquad} An interpolation-based quadrature
  method that works in two and three dimensions. The spectrally-accurate
  trapezoid rule is used in points in $\mathtt{Far}(\Gamma_i)$, while
  one dimensional interpolant is used for points in
  $\mathtt{Near}(\Gamma_i)$. $h$ is the arclength spacing.}
\end{wrapfigure}

In experiments, it is known that interactions between bounding walls and
granules cannot be ignored~\cite{qua-gan-you2021,
KaouiHartingMisbah2011_PRE, her-sto-gra2005}. Therefore, the necessary
modification to the code will be performed so that confinement can be
considered. This modification requires introducing an additional layer
potential on the bounding walls, and elements of this modification have
already been incorporated (Figure~\ref{fig:stenosis}). However, to
consider rheological properties of the suspensions, a periodic domain,
such as a Couette device, will be considered. This multiply-connected
domain requires additional modifications. PI Quaife has considered
suspensions with confinement for several projects including dynamic
suspensions~\cite{qua-bir2014, bys-sha-qua2020} and porous
media~\cite{chi-moo-qua2020, dea-qua-bir-jua2018, moo-che-chi-qua2022,
qua-gan-you2021}.

When granules are close to one another, which is inevitable in dense and
confined suspensions, the resulting integrals are nearly-singular. This
issue is well-known in the IE community and several algorithms have been
developed in the last 30 years, including panel-based
methods~\cite{helsing_2008a, helsing_tutorial_2012}, Quadrature by
Expansions (QBX)~\cite{din-hua-mar-tan2021, klo-bar-gre-one2013,
ste-bar2022, af2018sisc}, Barycentric-type formulae for Cauchy
integrals~\cite{bar2014, bar-wu-vee2015, ioa-pap-per1991}, and
interpolation-based methods~\cite{ying_2006, qua-bir2014,
rah-vee-zor-bir2015} (Figure~\ref{fig:interpquad}). The proposed
research, which is described next, will combine several methods
including singularity subtraction, analytic function theory, and
spherical basis representations.

@@

% ----------------------------------------------------------------------
\subsubsection{Incorporating confinement}
\label{subsec:Confinement}
